\vspace{1.0cm} \Large \bf Integrating influenza antigenic dynamics with molecular evolution

### 2013-11-14 response.tex

\date{}

\begin{document}

\maketitle

Original reviewer criticisms are in plain text.  Our responses follow in \textbf{bold}.  A detailed list of every criticism and exactly what was revised in the manuscript in response to this criticism is available at \url{https://github.com/trvrb/flux/issues?state=closed}.

@@

The one exception to the aforementioned praise on data availability is the influenza incidence data - I could not find any tables providing these data. They need to be provided.

\textbf{We apologize for this oversight.  The incidence data table used in our analysis has now been included with the Dryad archive and a link to this archive included under availability.  We've also included a reference to the BEAST tutorials on implementing these cartographic models.}

The authors need to define their nomenclature regarding clade / lineage / type / strain. For instance, they say: ``We find that 1 unit of antigenic drift within a clade increases incidence by 1.26 standard deviations, while 1 unit of drift in a clade of the same type decreases incidence by 0.61 standard deviations and 1 unit of drift in clades of differing type decreases incidence by 0.32 standard deviations. The term for across lineage Bt is small in magnitude.'' I had to read this many times before concluding that type refers to A or B, clade to refers to any of the four classifications. I am still not sure if lineage has a specific meaning or is simply a redundant word for clade.

\textbf{We've cleaned up our language regarding clades and lineages in the text.  We consider A/H3N2, A/H1N1, B/Vic and B/Yam to each represent separate \textit{lineages} of influenza.  We no longer refer to separate \textit{clades}.}

@@

Understanding antigenic drift of influenza in quantitative terms is clearly of great practical importance, and also an interesting intellectual problem in its own right. And it is nice to see antigenic maps for so many viral clades. However, I do not feel that the technical advances in ``mapping'' strains introduced here provide us much more confidence in the antigenic locations of strains than prior techniques do. Indeed, the authors themselves note (page 3) that the average predictive error of this new methods is comparable to that of Smith et all (2004). What might be more valuable, perhaps, are the results relating inferred antigenic drift and influenza incidence each season, by clade. But these results are quite possibly driven by sampling biases (see below), and I am not really convinced of their validity -- at least not without serious efforts to remove such sampling biases. Finally, the discussion of a ``diffusion'' model of drift was extremely confusing to me, and the fitted diffusion coefficients do not have a clear meaning or interpretation.

The authors do not seem to discuss this a problematic aspect of their results relating inferred drift to influenza disease incidence -- namely that biased sampling of clades, in each season, could cause the observed correlations. If clades were sampled for HI assays unequally in each season, as I suspect, then this fact alone might cause the trend towards apparently more drift in the clade that was sampled most intensively in each season (and to which most ILI was attributed). This is an obvious possible confound for the analysis presented in Fig. 4, and it should be ruled out by simulation of constant true rates of antigenic drift, variable sampling proportions in each season, and reconstructed drift using BMDS.

\textbf{We tested to see if there might be an bias in the estimates of year-to-year antigenic drift based on number of virus isolates collected by correlating correlating drift between years $i$ and $i-1$ against the ratio of the number of isolates from year $i$ to the number of isolates from year $i-1$.  We found no significant correlation (Pearson's $r = -0.01$, $p = 0.411$), suggesting our results to be unbiased with regard to sample count.  We have included this analysis in the manuscript.}

@@

Likewise, in Fig. 5, the negative correlation between drift in different clades in each season could, again, be driven by variation in the number of strains per clade that were subjected to HI each season. The authors could rule this out either by simulation or, more simply, be repeating their analysis after subsampling an equal number of isolates per clade each season for inclusion into the BMDS.

\textbf{The above analysis applies to this correlation as well.}

### 2013-11-14 flux.tex
Fix #22 and fix #23.
No correlation between sample counts and estimates of year-to-year drift.

We sought to summarize year-to-year patterns of antigenic drift by calculating the difference in mean location between consecutive years (Figure~\ref{jumps}A).
We estimate year-to-year antigenic drift for years 1992 to 2011 by calculating the average location along dimension 1 of phylogenetic lineages present in the tree at year $i$ and comparing this location to the average location of phylogenetic lineages present in the tree at year $i-1$.
There may often be large discontinuities in virus locations across the population; our use of difference in mean location is meant to capture both the distance between antigenic clusters and also the change in cluster frequency over consecutive years.
We observe greater heterogeneity in year-to-year antigenic drift in type A than in type B lineages (Figure~\ref{jumps}B), with standard deviation of year-to-year antigenic drift equal to 0.97 units in A/H3N2, 0.66 units in A/H1N1, 0.46 units in B/Vic and 0.26 units in B/Yam.
This analysis classifies drift only to the level of consecutive years; some coarse-graining of the timings of transition events will necessarily occur.

>>>>

We sought to summarize year-to-year patterns of antigenic drift by calculating the difference in mean location between consecutive years (Figure~\ref{jumps}A).
We estimate year-to-year antigenic drift for years 1992 to 2011 by calculating the average location along dimension 1 of phylogenetic lineages present in the tree at year $i$ and comparing this location to the average location of phylogenetic lineages present in the tree at year $i-1$.
There may often be large discontinuities in virus locations across the population; our use of difference in mean location is meant to capture both the distance between antigenic clusters and also the change in cluster frequency over consecutive years.
We observe greater heterogeneity in year-to-year antigenic drift in type A than in type B lineages (Figure~\ref{jumps}B), with standard deviation of year-to-year antigenic drift equal to 0.97 units in A/H3N2, 0.66 units in A/H1N1, 0.46 units in B/Vic and 0.26 units in B/Yam.
This analysis classifies drift only to the level of consecutive years; some coarse-graining of the timings of transition events will necessarily occur.
We tested to see if there could be an artifactual effect on the measurement of year-to-year drift based on the number of viral isolates collected by correlating drift between years $i$ and $i-1$ against the ratio of the number of isolates from year $i$ to the number of isolates from year $i-1$.
We found no significant correlation (Pearson's $r = -0.01$, $p = 0.411$), suggesting our results to be unbiased with regard to sample count.
